A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-06 podc_paper/algorithm-2.tex
Editing GC

\section{Garbage Collection}

\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTSINGLE}{\com\ #1}}
\begin{figure}
%\here{Does \var{leaf} have to be shared for helping when doing GC?}
\begin{algorithmic}[1]
\setcounter{ALG@line}{1}
\Statex \linecomment{Shared variable}
\begin{itemize}
\item \typ{Node} \var{root} \Comment{root of a binary tree of \tt{Node}s with one leaf for each process} % TODO: indent these lines in order to be more readable. I tries some times and i could not figure out doing that in algorithm.
\item \gc{\typ{*Node[]} \var{leafOfProcess}$[1..p]$ \Comment{Pointer to the leaf assigned to each process.}}
\item \gc{\typ{int[]} \var{last}$[1..p]$ \Comment{index of last block each process has dequeued an element from}}
\end{itemize}

\vspace{0.8em} %TODO: clean code

\Statex \linecomment{Thread-local variable}
\begin{itemize}
\item \typ{Node} \var{leaf} \Comment{process's leaf in the tree}
\end{itemize}

\vspace{0.4em} %TODO: clean code

\Statex $\blacktriangleright$ \typ{Node}
\begin{itemize}
\item \typ{Node} \fld{left}, \fld{right}, \fld{parent} \Comment{children and parent pointers initialized  when creating the tree}
\item \typ{Block}[0..$\infty$] \fld{blocks} \Comment{initially \var{blocks}[0] contains an empty block with all integer fields equal to 0}
\gc{\item \typ{PBRT<Block, int>} \fld{blocks} \Comment{PBRT of blocks each node contains a block and an index key}}
\item \typ{int} \head \Comment{position to attempt appending next \block\ to \fld{blocks}, initially 1}
\end{itemize}

\vspace{0.8em} %TODO: clean code

\Statex $\blacktriangleright$ \typ{Block}

\begin{itemize}
  	\item \typ{int} \fld{sum\sub{enq}, sum\sub{deq}}
  		\Comment{number of enqueue, dequeue operations in \fld{blocks} array up to this block (inclusive)}
  	\item \typ{int} \fld{super}
  		\Comment{approximate index of the block's superblock in \fld{parent.blocks}}
	\item[\com] Blocks in internal \nodes\ have the following additional fields
	\item \typ{int} \eleft, \eright
  		\Comment{index of last direct subblock in the left and right child}
  	\item[\com] Blocks in leaf \nodes\ have the following additional field
  	\item \typ{Object} \fld{element}
  		\Comment{if the block's operation is \opa{Enqueue}{x} then \fld{element}=\var{x}, otherwise \fld{element}=\nl}
  	\gc{\item \typ{Object} \fld{response}
  		\Comment{if the block's operation is \opa{Dequeue}{} then \fld{response} is the response of the\\ operation if it has been helped.}}
	\item[\com] Blocks in the root \node\ have the following additional field
	\item \typ{int} \size%
  		\Comment{size of the queue after performing all operations up to the end of this block}
\end{itemize}

%\Statex {\com\  Blocks in internal nodes have the following additional fields}
%\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in internal nodes}
%\begin{itemize}
%	\item \tt{int} \eleft, \eright
%  		\Comment{index of last direct subblock in the left and right child}
%\end{itemize}

%\Statex {\com\ Blocks in leaf nodes have the following additional field}
%\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}} \sf{\com\ the following additional field is used only for blocks in leaves}
%\begin{itemize}
%  \item \tt{\sl{Object} element}
%  \Comment{if the block's operation is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}
%\end{itemize}

%\Statex {\com\ Blocks in the root node have the following additional field}
%\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}} \sf{\com\ the following additional field is used only for blocks in the root}
%\begin{itemize}
%  \item \tt{\sl{int} \size}%
%  \Comment{size of the queue after performing all operations up to the end of this block}
%\end{itemize}

\end{algorithmic}
\caption{Objects used in tournament tree data structure \label{object-fields}}
\end{figure}

\subsection{Details of the Implementation}

We now discuss the queue implementation in more detail.  Pseudocode is provided in Figure \ref{pseudocode1} and \ref{pseudocode2}.
%We use $v.\var{blocks[i].num\sub{enq}}$ as shorthand for 
%$v.\var{blocks[i].sum\sub{enq} - blocks[i-1].sum\sub{enq}}$, that is, 
%the number of enqueues in the block.  (For $\var{i}=0$, $v.\var{blocks[0].num\sub{enq}} = 0$.)
%We use \var{num\sub{deq}} similarly.
%\here{Check if the num abbreviation is really needed in the code--how many times do we use it?}

@@

In our algorithm an \nf{Enqueue} or a \nf{Dequeue} remains in the \nf{blocks} array in the tree nodes even after they terminate. This makes the space used by the algorithm factor of the number of operations of invoked on the queue. Here in this section, we want to add a mechanism to free the memory allocated by the operations that are no longer needed and make the overall space used polynomial of $p+q$.

It is a common way to do garbage collection in batches. If we garbage collect the unnecessary blocks in the nodes every $p^2$ block appended to the root, the garbage collection cost is amortized over $p^2$ blocks which is $O(p^3)$ and $\Omega(p^2)$ operations.

>>>>

\section{Detailed Description of the Bounded-Space Implementation}
\label{reducing-details}

Here, we give more details of the bounded-space construction sketched in Section \ref{reducing}.
We first describe the modifications to the implementation.
It uses two additional shared arrays.
\begin{itemize}
\item \typ{Node[]} \var{leaf}$[1..p]$ \Comment{\var{leaf}$[k]$ is the leaf assigned to process $k$}
\item \typ{int[]} \var{last}$[1..p]$ \Comment{\var{last}$[k]$ indexes last block in $root$ that process $k$ has dequeued from}
\end{itemize}
The \fld{blocks} field is implemented as a pointer to the root of a RBT of \typ{Blocks} rather than an infinite array.
Each \typ{Block} has an additional field.
\begin{itemize}
\item \typ{int} \var{index} \Comment{position this block would have in the \fld{blocks} array}
\end{itemize}
To facilitate helping, each \typ{Block} in a leaf has one more additional field.
\begin{itemize}
\item \typ{Object} \var{response} \Comment{response of the operation in the block, if it is a dequeue}
\end{itemize}

@@

If we garbage collect the unnecessary blocks in the nodes every $p^2$ block appended to the root, the garbage collection cost is amortized over $p^2$ blocks which is $O(p^3)$ and $\Omega(p^2)$ operations.

### 2023-01-06 podc_paper/boundedspace.tex
Editing GC

We now consider how to modify our implementation to bound the space it uses.
We replace the \fld{blocks} array in each node by a red-black tree (RBT)
that stores the blocks.
Each block has an additional \fld{index} field
that represents its position within the original \fld{blocks} array, and
blocks in a RBT are sorted by \fld{index}.
Accessing the block in entry $i$ in the \fld{blocks} array can then be 
replaced by searching the RBT for the block with index $i$.
The attempt to append a new block in entry $i$ of the \fld{blocks} array on line \ref{cas}
is replaced by an attempt to insert a new block with index $i$ into the RBT.

@@

>>>>

In our implementation, an operation remains in the \nf{blocks} arrays forever. 
Thus, the space used is proportional to the number of operations that have been invoked.
Now, we sketch a mechanism to remove blocks that are no longer needed to use space
polynomial in $p$ and $q$, while maintaining polylogarithmic (amortized) step complexity.  For lack of space, details are in Appendix~\ref{reducing-details}.

We replace the \fld{blocks} array in each node by a red-black tree (RBT)
that stores the blocks.
Each block has an additional \fld{index} field
that represents its position within the original \fld{blocks} array, and
blocks in a RBT are sorted by \fld{index}.
Accessing the block in entry $i$ in the \fld{blocks} array can then be 
replaced by searching the RBT for the block with index $i$.
The attempt to append a new block in entry $i$ of the \fld{blocks} array on line \ref{cas}
is replaced by an attempt to insert a new block with index $i$ into the RBT.

@@

Each process takes $\tilde{O}(p)$ to run GC.\footnote{We use $\tilde{O}$ notation
to omit polylogarithmic factors; see Appendix \ref{reducing-details} for precise bounds.}
Since all processes may run GC,
a total of $\tilde{O}(p^2)$ steps is spent on a GC phase.
Between two of these GC phases, there are at least $p^2$ operations, so
GC contributes only $\tilde{O}(1)$ steps per operation in an amortized sense.
